{\chapter{A C K N O W L E D G M E N T S} } {\label{A C K N O W L E D G M E N T S} }{\par}{\textit{	} } {\par}{\par} {\textbf{\textit{	} } } {\par} 
	Much of this book originated in literary periodicals and magazines. Were it not for editors Alex Star, Paul Laity, Marykay Wilmer's, Paul Meyers cough, Adam Schatz, John Panatella, and Jackson Lear's, I never would have written about the right. It’s often assumed that academics who publish in nonacademic venues are trotting out their scholarly research for popular consumption, simplifying complex ideas first worked out in the laboratories of academe. For me, the process of writing this book has been the reverse: conservatism became a scholarly interest of mine through my nonacademic writing, and most of my ideas about the right were formulated in conversation with and writing for these editors, especially Alex and John.{\par} 	Intellectually, this book owes its inspiration to Arno Mayer and Karen Orren. No two scholars have done more to advance my under-standing of “the persistence of the old regime”—in Europe and the United States—than Karen and Arno. Against the conventional wisdom of the left and the right, which assumes that medievalism has been washed away by modernity, Karen and Arno opened my eyes to the “belated feudalism” of our post-feudal world. While they undoubtedly would disagree with my interpretation of conservatism, I could not have come to it without their enormously generative work.{\par} 	In the course of writing and revising these essays, I have been sustained by a broad circle of readers: historians and political scientists, poets and essayists, theorists and philosophers, literary critics and sociologists, journalists and editors. For their contributions to one or more of these essays, I would like to thank Jed Bahamian, Bruce Ackerman, Joel Allen, Gaston Alonso, Joyce Appleby, Moustafa Bayoumi, Sale Benjamin, Mar-shall Berman, Sara Bechtel, Keel Milgram, Norman Birnbaum, Steve Bonner, Dan Brook, Sebastian Budget, Josh Cohen, Peter Cole, Paisley Hurrah, Lizzie Donahue, Jay Driscoll, Tom Dump, John Dunn, Sam Farmer, Liza Feather stone, Jason Frank, Steve Fraser, Josh Freeman, Paul Framer, Sam Goldman, Many Goswami, Alex Gourevitch, Pete Hallward, Harry Harootunian, Chris Hayes, Doug Heywood, Dick Howard, David Hughes, Judy Hughes, Allen Hunter, Jack Jacobs, Ira Katz nelson, Gordon Later, Jill Lenore, Penny Lewis, Joe Lowndes, Steven Lukes, Kinko Matheson, Kevin Matt son, John Madeiras, Kathy New-man, Molly Nolan, Anne Norton, Jolie Alcott, Christian Parents, Di Peyton, Rick Perl stein, ROS Patches, Kim Phillipsfan, Kathy Polling, Aziz Reyna, Andy Rich, Andrew Ross, Kristin Ross, Saskia Lassen, Ellen Checker, George Scalable, Richard Seymour, Nikhil Singh, Quentin Skinner, Jim Sleeper, Rogers Smith, Katrina van den Level, John Wallach, Eve Wanna, Keith Whittington, Daniel Wilkinson, Wesley Yang, Brian Young, and Marilyn Young.{\par} A good portion of this material has been presented in work-shops and talks at universities across the country. I am grateful for the comments and suggestions I received on those occasions from Arash Abide, Anthony Appear, Band Barge, Sale Benjamin, Keel Milgram, Elizabeth Cohen, Josh Cohen, Julie Cooper, the late Jack Digging, Matt Evans, Nancy Fraser, Mark Grader, Nan Methane, Steve Macedon, Karina Mantegna, Andrew March, Tom Meet, Andrew Murphy, Andrew Norris, Anne Norton, Joshua Over, Philip Pet tit, Andy Polska, Robert Reich, Austin Sarah, Peter i would like to thank the following institutions for providing much needed release time from my teaching: the American Council of Learned Societies; the Princeton University Center for Human Values; the Off ice of the Provost at Brooklyn College; and the Professional Staff Congress of the City University of New York. An extra special thanks goes to my kitchen cabinet of first readers: Greg Gran din, Adana Hoff man, Robert Parkinson, and Scott Saul; to Marco Roth, who came up with the book’s title; to Charles Petersen, copy editor extraordinaire; to my students at Brooklyn College and the CUNY Graduate from Center, who have worked with me through the texts and tomes of the right; to Alexandra Hauler and Marc Schneider at Oxford University Press (OUT); and to David Mc bride, my editor at OUT, an unfailing source of excellent advice who believed in this project from its inception and shepherded it through with what seems to be effed ortless wisdom, patience, and grace.{\par} My greatest thanks go to Laura Brahma, who listened to these ideas when they were half-sentences and read them when they were half-baked. She has brought to these essays an eye for what matters and an unerring sense of taste. She is always and inevitably the only reader I want to please.{\par}